---
title: "Common Data Preparation in R with dplyr and tidyr"
author: Kuo Liu
date: \`r format(Sys.time(), "%B %d, %Y")`\
output: html_notebook
---

This R notebook serves as a companion for the Jupyter Notebook: _Common Data Preparation in Pandas_. Most examples in **Part 1** of the notebook are based on this [dplyr tutorial](http://genomicsclass.github.io/book/pages/dplyr_tutorial.html) from the Data Analysis for Genomics course by HardvardX.  

In **part 1** we study the 6 verbs in `dplyr`, i.e.:

- `select`: Select columns from a data frame.
- `filter`: Select rows (based on certain conditions) from a data frame.
- `arrange`: Reorder rows of a data frame.
- `mutate`: Modify existing columns or creat new columns in a data frame.
- `summarise`: A vector to single number operation that is usually used together with `groupby` to summarize results in each group. 
- `join`: Join two data frames by a key (or several keys).


In **Part 2**, we focus on 4 functions in `tidyr`.

- `gather`: Reshape a data frame from _wide to long_ and make certain column names themselves as new variables in rows. 
- `spread`: Reshape a data frame from _long to wide_ and make certain values in rows as a new column name. The inverse of `gather`
- `separate`: Split one column in a data frame into multiple new columns. 
- `unite`: Concatenate multiple columns into one column in a data frame. The inverse of `separate`.

We will also introduce the very convenient pipe operator `%>%`. 

## Part 1: Verbs in dplyr
Load library _dplyr_ 
```{r}
library(dplyr)

my_url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/msleep_ggplot2.csv"
df <- read.csv(url(my_url))
df <- as_tibble(df)
df
```


### Select columns
```{r}
# Approach 1: select by column name
select(df, name, sleep_total)

# Approach 2: select by column index
select(msleep, c(1,6))
```

### Select rows
There are 2 ways to select rows in `dplyr`

- Select rows by index: `slice`
- Select rows by condition: `filter`

Select rows by index.
```{r}
slice(msleep, 6:10)
```


Select rows with 1 condition:
```{r}
filter(df, sleep_total >= 16)
```

Select rows with multiple conditions:
```{r}
filter(df, sleep_total >= 3, bodywt > 1 | brainwt > 0.005)
```

As you can see from the examples above, in `dplyr`, usually the data preparation process is like this:

- First, you think about what data operations you want to perform
- Second, you find the right "verb" in `dplyr`
- Third, you provide the minimum needed information (function parameters) to accomplish the operation.

While in pandas, usually the thought process is not as consistent as in `dplyr`. Other than the operation itself, you need to think about the underline data structure a lot. 


### Reorder rows, using a pipe
```{r}
df %>% arrange(order)
```

Modify an existing column
```{r}
df %>%
  mutate(sleep_total = sleep_total / 24)
```


Create a new column
```{r}
df %>%
  mutate(rem_proportion = sleep_rem / sleep_total) %>%
  head
```


### Summarise data (within groups)
```{r}
df %>%
  group_by(order) %>%
  summarise(avg_sleep = mean(sleep_total),
            min_sleep = min(sleep_total),
            max_sleep = max(sleep_total),
            total = n())
```


### Join two data.frames 
```{r}
band_members
band_instruments

band_members %>% inner_join(band_instruments)
band_members %>% left_join(band_instruments)
band_members %>% right_join(band_instruments)
band_members %>% full_join(band_instruments)
```


## Part 2: 4 Verbs in tidyr
### Reshape data (wide-to-long and long-to-wide)
```{r}
library(tidyr)

# Gather  ======
df1 <- read.csv("supplies.csv")
df1 <- as_tibble(df1)
df2 <- df1 %>% gather(Name, Value, January:December )
df2

# Spread  =======
df2 %>% spread(key = Name, value = Value) %>% select(names(df1))
```


### Unite and separate columns
```{r}
df2 <- df1 %>% unite(Prod_cate, Product, Category)
df2 %>% separate(Prod_cate, c("Product", "Category"), sep = '_')
```

